\fontsize{18}{22}\selectfont A comprehensive study of the phylodynamics of SARS-CoV-2 in Europe

### 2021-03-24 reports/mt-thesis/results.tex
wip results mt

\textbf{The origin of SARS-CoV-2 in Europe} 
While the first case in Europe was reported to ECDC by France authorities on January 25, 2020, the first European wave is suspected to have started earlier but been undetected, mistaken for other common repiratory diseases in winter. In our model, we assume the origin of SARS-CoV-2 in Europe to be an imported case from China into any of the European countries. We analyze the time and destination of the first imported case in the infered epidemic trajectories and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI), Figure \ref{fig:firstEU} \todo{with final results}. The actual reported date to ECDC is in the tail of the $95\%$ credible interval. As shown in Figure \ref{fig:firstEUcountry} A, we place the destination country of this first imported case in Italy or France, with x of probability each.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
    \caption{First introductions of SARS-CoV-2. From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted. \textbf{A} Probability density of the time of first introduction for each deme. Each dotted line represents the first date when cases where reported to ECDC by deme color. In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1. For the other five demes, the distribution of the time of first migration into the deme is shown. \textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme. This first case corresponds to the first migration event from China to any of the European countries. \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
    \label{fig:firstEU}
    \todo[inline]{ticks for every month}
\end{figure}

Likewise, we analyze the first ten imported cases into Europe and obtain the probability distribution of the source and destination country, Figure \ref{fig:firstEUcountry} B and C.  The role of Italy as the receptor of these first cases is highlighted even more looking at several cases. Early introductions in Spain have a very low probabilty, looking like Spain was not a key player in the origin of SARS-CoV-2 in Europe. China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China. \todo{with final results}

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj04.png}
    \caption{First introductions of SARS-CoV-2. From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted. \textbf{A} Probability density of the time of first introduction for each deme. Each dotted line represents the first date when cases where reported to ECDC by deme color. In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1. For the other five demes, the distribution of the time of first migration into the deme is shown. \textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme. This first case corresponds to the first migration event from China to any of the European countries. \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
    \label{fig:firstEUcountry}
    \todo[inline]{ticks for every month}
\end{figure}

@@

Since in our model we consider the different European countries, we can also estimate from the infered epidemic trajectories the origin of the epidedmic in each particular country. In Figure we have the origin time for each country compared to the ddate of the first reporrted case to ECDC by that country. In average, we predict a delay in detection of x days for the Eurropean countries, out of the $95\% CCI$ for France, Germany and Italy. Origin in Spain close to the official detection day. We estimate that the epidemic in any of the European countries was seeded by an imported case from China as seen in Figurem with some small prrobability for Germany and Italy of having orriginatetdd from a case from each other, and mostly half of the rpobability in the case of Spain from the epidemdic started with a case imported from one of the others Euroopean countries. Same than before, we can look into the first 10 imported cases into each country and now we see that the transmission between european countries adquire much moer importance,  so once SARS-CoV-2 entered Europe, the transmission between european countrries were soon more important than the cases coming from China. This could relate also to the decrease in tthe trravel to and from China seen in figure.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj05.png}
    \caption{First introductions of SARS-CoV-2. From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted. \textbf{A} Probability density of the time of first introduction for each deme. Each dotted line represents the first date when cases where reported to ECDC by deme color. In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1. For the other five demes, the distribution of the time of first migration into the deme is shown. \textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme. This first case corresponds to the first migration event from China to any of the European countries. \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
    \label{fig:firstEUcountry}
    \todo[inline]{ticks for every month}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj06.png}
    \caption{Countries ordered by the time of its first introduction, i.e. first case in the country. Each row is the order for one of the subsampled epidemic trajectories and each column represents the position relative to the other countries first introduction, e.g. in first position for all epidemic trajectories is China since it is the origin of the epidemic.} 
    \label{fig:first_imig}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj07.png}
    \caption{Order of countries by the time of its first migration out of the country, i.e. first exported case to other country. Each row is the order for one of the subsampled epidemic trajectories and each column represents the position relative to the other countries first migration, e.g. in first position for all epidemic trajectories is China since it was the first country with exported cases of SARS-CoV-2 to other regions.}
    \label{fig:first_omig}
    \todo[inline]{I don't like much these "order" plots but maybe they could be useful to detect interesting patterns?}
\end{figure}

@@

\begin{figure}[p]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
    \caption{First introductions of SARS-CoV-2. From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted. \textbf{A} Probability density of the time of first introduction for each deme. Each dotted line represents the first date when cases where reported to ECDC by deme color. In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1. For the other five demes, the distribution of the time of first migration into the deme is shown. \textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme. This first case corresponds to the first migration event from China to any of the European countries. \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
    \label{fig:first}
    \todo[inline]{ticks for every month}
\end{figure}

@@

\begin{figure}[p]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj07.png}
    \caption{First introductions of SARS-CoV-2. From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted. \textbf{A} Probability density of the time of first introduction for each deme. Each dotted line represents the first date when cases where reported to ECDC by deme color. In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1. For the other five demes, the distribution of the time of first migration into the deme is shown. \textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme. This first case corresponds to the first migration event from China to any of the European countries. \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
    \label{fig:first}
    \todo[inline]{ticks for every month}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{210205_europe10_figtraj01.png}
    \caption{Inferred population size summary statistics for each deme over time. The line represents the median population trajectory and the interval is the 95\%  credible interval in log scale from a random subsampled set of inferred epidemic trajectories.}
    \label{fig:gribbon}
    \todo[inline]{Specify that the random subset are 500 trajectories in all figure legends}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj02.png}
    \caption{Inferred epidemic trajectories over time. A random subsampled set of 500 trajectories is plotted. In each subplot, the trajectories (solid lines) are compared with the ECDC cumulative case count data (dashed line) in log scale.}
    \label{fig:trajs}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj11.png}
    \caption{5-days reporting rate, calculated as the cumulative number of ECDC reported cases by the median number of cumulative inferred cases in intervals of 5 days until 8th of March. }
    \label{fig:reported}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj12.png}
    \caption{First introductions of SARS-CoV-2. From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted. \textbf{A} Probability density of the time of first introduction for each deme. Each dotted line represents the first date when cases where reported to ECDC by deme color. In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1. For the other five demes, the distribution of the time of first migration into the deme is shown. \textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme. This first case corresponds to the first migration event from China to any of the European countries. \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
    \label{fig:first}
    \todo[inline]{ticks for every month}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj08.png}
    \caption{Median and 95\% credible interval for the cumulative number of events (within-region transmissions and migrations to the country (incoming) and from the country (outgoing) over time.}
    \label{fig:events}
    \todo[inline]{change legend outgoing migration}
\end{figure}

@@

  \begin{minipage}[t]{0.4\textwidth}
  \includegraphics[width=\textwidth]{210205_europe10_figtraj09a.png}
  \label{fig:migs1}
  \end{minipage}
  \begin{minipage}[t]{0.4\textwidth}
  \includegraphics[width=\textwidth]{210205_europe10_figtraj09b.png}
  \label{fig:migs2}
  \end{minipage}
  \begin{minipage}[t]{0.4\textwidth}
  \includegraphics[width=\textwidth]{210205_europe10_figtraj09c.png}
  \label{fig:migs3}
  \end{minipage}
  \caption{Migration flux among demes over the three periods defined in the analysis.}
  \label{fig:migs}
\end{figure}

@@

\subsubsection*{GLM predictors}
\todo[inline]{TODO}

Which channels were the main sources of transmission across national borders.

@@

Not sure if it is necessary. But maybe include, briefly, the values of estimate R0, migration and sampling rates?

@@

>>>>

\textbf{The origin of SARS-CoV-2 in Europe} 
While the first case in Europe was reported to ECDC by France authorities on January 25, the first European wave is suspected to have started earlier but been undetected, mistaken for other common repiratory diseases in winter. In our model, we assume the origin of SARS-CoV-2 in Europe to be an imported case from China to any of the European countries. We can select a single event of interest from the inferred epidemic trajectories, and summarise its time and location. To study the origin of SARS-CoV-2 in Europe, we analyze the time and destination of the first imported case and estimate the origin of the European epidemic on January 2020 ($95\%$ CCI), Figure \ref{fig:firstEU} \todo{with final results}. The actual reported date to ECDC is in the tail of the $95\%$ credible interval. As shown in Figure \ref{fig:firstEUcountry} A, we place the destination country of this first imported case in Italy or France, with x of probability each.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj01.png}
    \caption{\textbf{First case of SARS-CoV-2 in Europe.} In the histogram, we count how many of the inferred epidemic trajectories started the European epidemic each day. From these counts, we estimate the probability density of the origin in Europe (in grey). The median date is drawn with the solid vertical line, and the date of the first reported European case to ECDC with a dashed vertical line.}
    \label{fig:firstEU}
\end{figure}

Likewise, we analyze the first ten imported cases to Europe to estimate the main source of exported SARS-CoV-2 cases early on and their destination. China is the main source of these first introductions, supporting the hypothesis of the European epidemic seeded by several and not just one imported cases from China, Figure \ref{fig:firstEUcountry} B.  The role of Italy as the receptor of these first cases is highlighted even more looking at several cases. Early introductions in Spain have a very low probabilty, looking like Spain was not a key player in the origin of SARS-CoV-2 in Europe. \todo{with final results}  Figure \ref{fig:firstEUcountry} C.

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj02.png}
    \caption{\textbf{First imported cases of SARS-CoV-2 to Europe.} \textbf{A} We estimate the probability distribution of the time for the first imported case to Europe (in grey). Each row highlights the probability of this first case going to the indicated country (in color). \textbf{B} In a similar way, we considered the time of the ten first imported cases into any of the European country, and plot the time probability distribution (in grey). In the rows, we plot the probability of each country being the destination country of any of the first 10 imported cases to Europe (in color). \textbf{C} Same probability distribution of the time for the 10 first imported cases to any European country (in grey) than in B, but in each row we highlight in color the 
    probability of these cases coming from a specific country.}
    \label{fig:firstEUcountry}
    \todo[inline]{ticks for every month}
\end{figure}

@@

Since in our model we consider the different European countries, we can also estimate from the infered epidemic trajectories the origin of the epidedmic in each particular country. In Figure we have the origin time for each country compared to the ddate of the first reporrted case to ECDC by that country. In average, we predict a delay in detection of x days for the Eurropean countries, out of the $95\% CCI$ for France, Germany and Italy. Origin in Spain close to the official detection day. We estimate that the epidemic in any of the European countries was seeded by an imported case from China as seen in Figurem with some small prrobability for Germany and Italy of having orriginatetdd from a case from each other, and mostly half of the rpobability in the case of Spain from the epidemdic started with a case imported from one of the others Euroopean countries. Same than before, we can look to the first 10 imported cases to each country and now we see that the transmission between european countries adquire much moer importance,  so once SARS-CoV-2 entered Europe, the transmission between european countrries were soon more important than the cases coming from China. This could relate also to the decrease in tthe trravel to and from China seen in figure.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj03.png}
    \caption{\textbf{Country-level origin of SARS-CoV-2 epidemic.} For China, we consider the origin of the epidemic as the first local case while for the European countries we select the first imported case into the country. \textbf{A} We estimate the probability distribution of the date for the first case in each of the locations included in the analysis (in grey). We highlight the conditional probability for the source of the first case in each country (in color).  \textbf{B} The ten first imported cases into the country are selected, and we estimate the probability distribution of their occurrence. As in A, we show the conditional probability of the source country.}
    \label{fig:firstcountry}
\end{figure}

@@

@@

@@

\begin{figure}[p]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj04.png}
    \caption{\textbf{Timing of key first epidemic events.} For each country, we select the key epidemic events: first local transmission case, first death, recovery or isolated case (same event category for the birth-death model), first imported and first exported case. We summarise the timing of these events from the inferred trajectories and obtain a probability distribution for each of them. The first detection of SARS-CoV-2 in each country reported to ECDC is shown with a dashed line.}
    \label{fig:first_events}
\end{figure}

@@

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.8\textwidth]{210205_europe10_figtraj05.png}
    \caption{\textbf{Active SARS-CoV-2 infections by country from the origin of the pandemic to March 8, 2020}. We compute the median number and 95\%  credible interval of active SARS-CoV-2 infections for each analysis day by country. We plot the median value for the country with a coloured line and the 95\% credible interval of active infections with a shaded area.}
    \label{fig:gribbon}
\end{figure}

\begin{figure}[p]
    \centering
    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj06.png}
    \caption{\textbf{SARS-CoV-2 inferred cases and ECDC confirmed cases by country over time.} To compare with the official ECDC confirmed cases in a analogous time frame, we consider the no longer infectious cases inferred by the model that appear after a mean period of ten days post-infection. Each line represent the cumulative number of inferred cases in log scale for each of the 500 analysed epidemic trajectories with time in the x-axis. The ECDC confirmed case counts are shown with dashed lines.}
    \todo[figure]{Why is till March 20?}
    \label{fig:trajs}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.9\textwidth]{210205_europe10_figtraj07.png}
    \caption{\textbf{Weekly reporting proportion of new SARS-CoV-2 cases.} We compute the number of new confirmed cases by week reported to ECDC and divide it by the median number of inferred cases (no longer infectious cases) in the same week. For each week, the point represents the median reporting proportion, and the shaded area the 95\% credible interval.}
    \label{fig:reported}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{210205_europe10_figtraj08.png}
    \caption{\textbf{Evolution of the ratio of local transmissions by imported cases for the SARS-CoV-2 early epidemic in China, France, Germany, Italy, Spain and other European countries}. We compute for each day the ratio of new local infections by new imported cases for each of the analysed epidemic trajectories. Each line depicts the evolution of this ratio in one epidemic trajectory, coloured by the total number of estimated infections by the model. The black line is the median value for the ratio local/imported cases.}
    \label{fig:events}
    \todo[inline]{change legend outgoing migration}
\end{figure}

@@

  \begin{minipage}[t]{0.4\textwidth}
  \includegraphics[width=\textwidth]{210205_europe10_figtraj09a.png}
  \label{fig:migs1}
  \end{minipage}
  \begin{minipage}[t]{0.4\textwidth}
  \includegraphics[width=\textwidth]{210205_europe10_figtraj09b.png}
  \label{fig:migs2}
  \end{minipage}
  \begin{minipage}[t]{0.4\textwidth}
  \includegraphics[width=\textwidth]{210205_europe10_figtraj09c.png}
  \label{fig:migs3}
  \end{minipage}
  \caption{\textbf{International imported case dynamics by analysis constat migration rate periods.} One chord diagram is drawn for each of the main periods of the analysis defined by the change times of the migration rates in the model. Each flux represents the infections exported from one country to another during the full period. The median value of infections for each flux is indicated in the plot.}
  \label{fig:migs}
\end{figure}

@@

@@

Not sure if it is necessary. But maybe include, briefly, the values of estimate R0, migration and sampling rates?
Maybe a table in the appendix? Or the plots with the priors.

@@

\begin{figure}[p]
    \centering
    \includegraphics[width=\textwidth]{particles_comparison.png}
    \caption{First introductions of SARS-CoV-2. From the set of random subsampled trajectories, the first introduction time for each epidemic trajectory is recorded and the probability distribution over all these times is plotted. \textbf{A} Probability density of the time of first introduction for each deme. Each dotted line represents the first date when cases where reported to ECDC by deme color. In the case of China, the distribution of the origin time is plotted, since in the analysis we defined Chine as the origin of the epidemic with probability 1. For the other five demes, the distribution of the time of first migration into the deme is shown. \textbf{B} Stacked probability density of the destination of first introduction into Europe coloured by the destination deme. This first case corresponds to the first migration event from China to any of the European countries. \textbf{C} Stacked probability density of the source of the first introduction for each deme coloured by the source deme of the introduction.}
    \label{fig:particles}
    \todo[inline]{ticks for every month}
\end{figure}
